A Quantitative Genetic Linked Selection Model of the Negative Selection Process in Humans

### 2023-07-13 manuscript/manuscript.tex
ms updates

\begin{figure}[htbp] \centering
    \includegraphics[width=\textwidth]{figures/figure_4.pdf} 
    \caption{The distribution of fitness effects of new mutations estimates for
        Yoruba individuals. (A) The DFEs using sparse (left column) and
        full-coverage (right column) tracks, across different annotation models
        (row). Color indicates the feature type. (B) The DFE of the
        full-coverage CDS-priority model comparing the estimates across
    populations.}
  \label{fig:figure-3}
\end{figure}

@@

However, across all of our annotation models, new deleterious mutations in
conserved feature classes (CADD tracks, CDS, and PhastCons regions) were
consistently estimated to have strongly deleterious effects (Figure
\ref{fig:figure-4}A), consistent with previous work
\parencite{McVicker2009-ax,Murphy2022-sj}. The DFE estimates for CADD and
PhastCons regions consistently places $\ge 75\%$ of mass on the largest
selection coefficient, $s=10^{-2}$. The CADD 6\% DFE estimates imply an average
selection coefficient of $\bar{s} = 0.65\%$ for CEU, $\bar{s} = 0.57\%$ for
CHB, and $\bar{s} = 0.79\%$ for YRI. Similarly, the PhastCons Priority model
implies average selection coefficient estimates of $\bar{s} = 0.63\%$, $\bar{s}
= 0.58\%$, and $\bar{s}= 0.77\%$ for CEU, CHB, and YRI respectively for
PhastCons regions.

@@

However, across the strong selection grid runs,

However, we find that there is considerable uncertainty in how strong selection
is against the most deleterious mutations. For example, in the CADD 6\%
full-track model, new mutations are estimated to have a selection coefficients
of $s = 10^{-1}$ with 42\% chance, $s=10^{-2}$ with 7.8\% chance, and
$s=10^{-3}$ with 41\% chance. This creates a bimodal DFE (Figure
\ref{fig:figure-3}A, second row, left column), which is mirrored across the
other CADD models. Moreover, these DFE estimates suggest a very high average
selection coefficient across CADD 6\% regions of ${\bar{s}} \approx 0.043$ for
YRI.

Given that previous methods have typically not estimated the DFE up to
selection coefficients of $s=10^{-1}$, we ran additional ``constrained grid"
models using a DFE only up to $s=10^{-2}$. Across all populations, these
constrained grid models fit the data slightly less well than the default grid
by about one percentage point less in $R_\text{LOCO}^2$ (see Supplementary
Table \ref{supp:tbl-r2}). As expected, the constrained grid models distributed
mass onto the next strongest selection coefficient, $s=10^{-2}$, and the
average selection was reduced about six-fold to ${\bar{s}} \approx 0.0079$ for
YRI. Our constrained-grid DFE estimates for the Feature Priority model are
qualitatively consistent with the U-shaped DFEs found for amino acids through
Poisson Random Field method \parencite{Boyko2008-tj}. While the DFE may truly
be bimodal for highly-conserved regions and have a very high average selection
coefficient, a likely alternative is that these results reflect a
non-identifiability issue between selective effects of $s=10^{-1}$ and
$s=10^{-2}$. We will refer to this as the \emph{strong-selection
non-identifiability} issue.

We propose that one mechanism for this non-identifiability is that mutations
extremely harmful to fitness are difficult to distinguish from a smaller
effective population. Intuitively, as $s$ increases, the average persistence
time (in generations) of the mutation in the population goes below one
generation, i.e. death of the carrier of the \emph{de novo} mutation. In this
hypothetical limit, each occurrence of a very harmful mutation nearly acts like
an instantaneous death that in effect is just a smaller population size. One
way to test this hypothesis is to look to see if there is a systematic positive
relationship across models between average selection coefficient and $\pi_0$,
which is includes the drift-effective population size $N_e$.

We find this is the case for all of our CADD 6\% models. Across every
population, average selection was about $7.1$ times larger using the default
grid, and $\pi_0$ was $5.6\%$ larger. There was no similar consistent change in
mutation rate estimates among populations. In the CADD 6\% model, genome-wide
average reduction factor $\bar{B}$ was $\approx 6.1\%$ lower in the default
versus constrained grid. Overall, this suggests that the linked selection
signal alone cannot differentiate very strong selection from a slightly smaller
drift-effective population size.

Despite the strong-selection non-identifiability issue, we were curious how
stable these estimates are across different reference populations. Overall, the
DFE estimates were consistent over all populations and models (Supplementary
Materials Section \ref{supp:fits}). Here, we highlight the Feature Priority
model since though it is poor-fitting, the DFE estimates are more readily
interpretable since it is based on gene features. Figure \ref{fig:figure-3}B
shows consistent DFE estimates across populations, with narrow uncertainty as
estimated by a block jackknife procedure. Coding sequences have a bimodal DFE
consistent with the strong-selection non-identifiability issue. Additionally,
coding sequences have a very uncertain neutral mode in the DFE. This is
expected given the different level of selective constraint between synonymous
and non-synonymous sites.

Additionally, the PhastCons class of this Feature Priority model, which
contains highly conserved non-coding elements, is under strong selective
constraint as expected. Finally, we highlight one result from our PhastCons
Priority annotation model (Figure \ref{fig:figure-3}A bottom row): the DFE
estimate for coding sequences excluding PhastCons regions is estimated as
neutral. This too is expected; the selection signal in coding regions is
absorbed by the PhastCons feature, leaving only conditionally neutral sites.

>>>>

\begin{figure}[htbp] \centering
    \includegraphics[width=\textwidth]{figures/figure_4.pdf}

    \caption{The distribution of fitness effects of new mutations estimates for
        YRI reference samples. (A) The DFEs using sparse (left column) and
        full-coverage (right column) tracks, across different annotation models
        (row). Color indicates the feature type. (B) The DFE of the
    full-coverage Feature Priority model comparing the estimates across
populations. Although this model fit the data less well than alternatives, its
results are more interpretable.}

  \label{fig:figure-3}
\end{figure}

@@

However, across all of our annotation models, new deleterious mutations in
conserved feature classes (CADD tracks, CDS, and PhastCons regions) were
consistently estimated to have strongly deleterious effects (Figure
\ref{fig:figure-4}A), consistent with previous work
\parencite{McVicker2009-ax,Murphy2022-sj}. The DFE estimates for CADD and
PhastCons regions consistently places $\ge 75\%$ of mass on the largest
selection coefficient, $s=10^{-2}$. The CADD 6\% DFE estimates imply an average
selection coefficient of $\bar{s} = 0.65\%$ for CEU, $\bar{s} = 0.57\%$ for
CHB, and $\bar{s} = 0.79\%$ for YRI. Similarly, the PhastCons Priority model
implies average selection coefficient estimates of $\bar{s} = 0.63\%$, $\bar{s}
= 0.58\%$, and $\bar{s}= 0.77\%$ for CEU, CHB, and YRI respectively for
PhastCons regions. Our DFE estimates for the Feature Priority model are
qualitatively consistent with the U-shaped DFEs found for amino acids through
Poisson Random Field method \parencite{Boyko2008-tj}.

@@

However, across the strong selection grid runs, we find indications of model
non-identifiability. First, estimates of the DFE with the strong grid are
strangely bimodal (Supplementary Figure \ref{suppfig:strong-sel-dfe}). For
example, under the CADD 6\% strong selection grid model, new mutations are
estimated to have a selection coefficients of $s = 10^{-1}$ with 43\% chance,
$s=10^{-2}$ with 7.8\% chance, and $s=10^{-3}$ with 41\% chance. We propose
that one mechanism for this non-identifiability is that mutations extremely
harmful to fitness are difficult to distinguish from a smaller effective
population. Intuitively, as $s$ increases, the average persistence time (in
generations) of the mutation in the population goes below one generation, i.e.
death of the carrier of the \emph{de novo} mutation. In this hypothetical
limit, each occurrence of a very harmful mutation nearly acts like an
instantaneous death that in effect is just a smaller population size. One way
to test this hypothesis is to look to see if there is a systematic positive
relationship across models between average selection coefficient and $\pi_0$,
which is includes the drift-effective population size $N_e$.

We find this is the case for all of our CADD 6\% models. Across every
population, average selection was about $7.1$ times larger using the default
grid, and $\pi_0$ was $5.6\%$ larger (see Supplementary Materials Figure
\ref{suppfig:sel-ident}). There was no similar consistent change in mutation
rate estimates among populations. In the CADD 6\% model, genome-wide average
reduction factor $\bar{B}$ was $\approx 6.1\%$ lower in the default versus
constrained grid. Overall, this suggests that the linked selection signal alone
cannot differentiate very strong selection from a slightly smaller
drift-effective population size.

Given that it is debated how strongly demography impacts the deleterious
mutation load
\parencite{Torres2018-ni,Torres2020-hc,Lohmueller2008-qi,Simons2014-sv,Simons2016-cs},
we were curious how consistent our DFE estimates are across populations.
Overall, we find DFE estimates are relatively stable across populations and
models (Supplementary Materials Section \ref{supp:fits}). Only in our Feature
Priority model (Figure \ref{fig:figure-4}B, top row) do we see a slightly
different DFE estimate for coding sequences between YRI and CEU/CHB
populations, but this could be due to the poorer fit this model has to data.

Although the Feature Priority model fits the data less well than alternative
models, its features are more interpretable. We find that our B' method
estimates a bimodal DFE for coding sequences for the Feature Priority model,
with a large mass placed on $10^{-3} \le s \le 10^{-2}$ and another on the
neutral class $s=10^{-8}$. This is expected, given that the synonymous and
non-synonymous sites that constitute coding sequences are under vastly
different levels of constraint. Moreover, features expected to be only weakly
constrained such as introns and UTRs have the bulk of DFE mass on the neutral
class, with a small but significant amount of mass ($\approx 3\%$) placed on
$s=10^{-2}$. As expected, the DFE for PhastCons regions suggests it is under
strong selective constraint; however we note that block jackknife based
uncertainty estimates suggest the model is uncertain whether there is some mass
on the neutral class. Finally, we highlight one result from our PhastCons
Priority annotation model (Figure \ref{fig:figure-3}A bottom row): the DFE
estimate for coding sequences excluding PhastCons regions is estimated as
neutral. This too is expected; the selection signal in coding regions is
absorbed by the PhastCons feature, leaving only conditionally neutral sites.
